---
title: "Compare with code DES"
author: "Eline Krijkamp"
date: "2025-11-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Code we had from course
```{r cars}
discount_at <- function(amount, time_point, rate) {
  # This function can be used to discount an amount at a specific
  # point in time.
  #
  # IMPORTANT! The input arguments 'time_point' and 'rate' need
  # to be defined in the same time unit (i.e., years or days)!!!
  #
  # INPUT:
  # - amount            the amount to be discounted
  # - time_point        the time point at which the amount is to be discounted
  # - rate              the discount rate (% per year)
  #
  # USES:
  #
  # OUTPUT:
  # - out
  #   [1] disc_amount   real      the discounted amount
  # Calculate the discount factor at the time point of interest
  disc_factor <- 1 / ((1 + (rate / 100)) ^ time_point);
  # Calculate the discounted amount
  disc_amount <- amount * disc_factor;
  # Define output object
  out <- c(disc_amount);
  return(out)
}
```

Test with an example about what the present value would be for a fix amount 
```{r}
# Discount $100 received in 3 years at 3% per year
amount <- 100
time_point <- 3
rate <- 3

pv_single <- discount_at(amount, time_point, rate)
cat("Present Value of $", amount, "in", time_point, "years at", rate, "% =", round(pv_single, 2), "\n")



```

Formula assuming that over a time period the same amount is accumulated each period 
```{r}

discount_fromto <- function(amount, time_from, time_to, time_step, rate) {
  # This function can be used to discount an amount over a specific
  # time interval in steps.
  #
  # IMPORTANT! The input arguments 'time_from', 'time_to', 'time_step' 
  # and 'rate' need to be defined in the same time unit (i.e., years 
  # or days)!!!
  #
  # INPUT:
  # - amount                the amount to be discounted
  # - time_from             the time point at which timing sequence is to be started
  # - time_to               the time point at which timing sequence is to be stopped
  # - time_step             the interval of the timing sequence
  # - rate                  the discount rate (% per year)
  #
  # USES:
  #
  # OUTPUT:
  # - out
  #   [1] cum_disc_amount   real      the sum of the discounted amounts
  # Define the time sequence of interest
  time_seq <- seq(from = time_from, to = time_to, by = time_step);
  # Calculate the discount factor at the time point of interest
  disc_factors <- 1 / ((1 + (rate / 100)) ^ time_seq);
  # Calculate the discounted amounts
  disc_amount <- amount * disc_factors;
  # Calculate the sum of the discounted amounts
  cum_disc_amount <- sum(disc_amount);
  # Define output object
  out <- c(cum_disc_amount);
  return(out)
}

```

Voorbeeld 

```{r}

# Suppose we receive $100 at the end of each year for 5 years.
# We’ll find the total present value of these payments.

amount <- 100
time_from <- 0
time_to <- 3
time_step <- 1
rate <- 3

pv_series <- discount_fromto(amount, time_from, time_to, time_step, rate)
cat("Cumulative Present Value of $", amount, "/year for", time_to, "years at", rate, "% =", round(pv_series, 2), "\n")
```

[OBSERVATION]: als je van 0 naar 3 gaat, heb je het dus over 4 jaar (0, 1, 2, 3). Want bij rate = 0, krijg je 400 euro. En dus heb je 4 x 100 euro. In de beschrijving is het dus essentieel om te vermelden hoe er moet worden om gegaan met time_from and time_to. 

```{r}
 time_seq <- seq(from = time_from, to = time_to, by = time_step);
time_seq
```



Volgens de beschrijving die code aanpassen. Nu deden we alles jaarlijks, maar stel je wilt naar maandelijk. Dan zegt de code dat je 
"  # IMPORTANT! The input arguments 'time_from', 'time_to', 'time_step' # and 'rate' need to be defined in the same time unit (i.e., years  or days)!!!" 

Dus dat interpreren wij als

```{r}
amount <- 100/12 # adjust to monthly (although not specified in the description )
time_from <- 0
time_to <- 3*12 # adjust to months 
time_step <- 1/12 # indicate months
rate <- 3/12 # adjust a annual rate to monthly 

pv_series <- discount_fromto(amount, time_from, time_to, time_step, rate)
cat("Cumulative Present Value of $", amount, "/months for", time_to, "months at", rate, "% =", round(pv_series, 2), "\n")



```
Dit geeft echt een verkeerde uitkomst. 

Waarschijnlijk moet de beschrijving als volgt:
 # IMPORTANT! The input arguments 'time_from' and 'time_to' as well as "rate" need to be on the annual scale. 'time_step' and `amount` need to be in the same unit as the simulation (i.e., years  or days)!!!" 


```{r}
amount <- 100/12
time_from <- 0
time_to <- 3
time_step <- 1/12
rate <- 3/12

pv_series <- discount_fromto(amount, time_from, time_to, time_step, rate)
cat("Cumulative Present Value of $", amount, "/months for", time_to, "months at", rate, "% =", round(pv_series, 2), "\n")


```
