---
title: "Reference prices: lookups by domain/category/unit & years (table vs single value)"
author: "tatooheene team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Reference prices: filters, years, currency, and single-value output}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7, fig.height = 4.2,
  dpi = 120
)
library(tatooheene)
library(dplyr)
library(tidyr)
library(lifecycle)
```

## What this vignette covers

-   Where the reference prices live: tatooheene::df_ref_prices.

-   How to use nl_ref_prices() to:

    -   filter by **domain, category, unit,**

    -   select **one or multiple years** (e.g., 2022 or c (2022, 2023) ),

    -   choose **currency** ("EUR" or "INT\$"; PPP adjusted), and

    -   return either a **table** or a **single numeric value** (with a printed label).

> The function is `r lifecycle::badge("experimental")`. Column names reflect the CBS table embedded in `df_ref_prices`.

## The underlying data

```{r peek-data}
df_ref_prices %>%
  head()
```

## Quick starts

### 1) Default: all filters = "all" -\> table of all reference prices over all years

```{r default}
nl_ref_prices()
```

### 2) Specific **year**\* & **unit** -\> single value

```{r specific}
nl_ref_prices(year = 2022, unit = "Emergency care")
```

### 3) Years values for a range of years

```{r range}
nl_ref_prices(year = 2018:2020, category = "Nursing")
```

### 4) Multiple selections return multiple columns

```{r multiple}
nl_ref_prices(
  year     = 2022:2024,
  category = "Nursing",
  unit     = c("Nursing day excluding personnel costs, hospital",
               "Nursing day including personnel costs, hospital")
)
```

## Error messages you might see

-   **"Year out of range …"** You asked for a year outside the available year columns in `df_ref_prices`. Use a valid year or pass `"all"` for all years.

-   **"For output = "value_label", provide exactly one year and a single (category, unit) combination."** Single value mode needs exactly one year **and** one (`category`, `unit`) (and, if used, a single domain) so that exactly one row remains.
